Bioinformatics analysis of bisulfite-sequencing data

### 2014-11-22 Rmarkdown/a_statistical_model_of_methlyC-seq.Rmd
Refactor webs analysis and stat model chapters

\section{Notes from outline.md (plus other notes added along the way)}

* Chapter Overview
* Notation
  * See `notation.Rmd`
* Estimating $Z$, $M$, $U$ and $\beta$
  * See `notation.Rmd`
  * Current approaches
* Statistical properties of $z$, $m$, $u$ and $\beta$
  * See `notation.Rmd`
  * Marginal distributions
  * Dependence/correlation over $i$ (see also __Chapter X__)
    * Li _et al._ compute $\beta$ values and compute correlations within a sample as a function of genomic distance.
    * Arabidopsis compute $\beta$ values and compute correlations within a sample as a function of genomic distance.
    * Lister (2009) Fig. 3 and Sup. Fig. 9
    * Lister _et al._ compute both within-read and across-read correlations or dependencies of DNA methylation.
    * Akulenko, R., _et al._ compute gene-level $\beta$ values and compute the correlations between pairs of genes across samples as a function of genomic distance.
  * Dependence/correlation over $j$ (?)
    * Akulenko, R., _et al._ compute gene-level $\beta$ values and compute the correlations between pairs of genes across samples as a function of genomic distance.

@@

To begin, I consider the simple experiment of performing methylC-seq on a single sample. Once I have described the framework for a single sample, I extend it to multiple samples. The ideas here are simple, although when extended to their full generality the notation becomes messy.

@@

Even with only a single sample, there are several levels of variation to consider, which have a hierarchical structure. I separate this structure into two stages, as illustrated in __CARTOON FIGURE OF METHYLC-SEQ EXPERIMENT__:

@@

__UP TO HERE__

$\beta$-values are the _de facto_ standard unit for reporting methylation levels due to their natural interpretation as the average level of methylation at the locus. However, they are not necessarily the best unit for statistical inference. This is because a $\beta$-value is an estimate of a proportion and there are a well-known challenges when working with proportion data, such as:

1. Proportions are bound between 0 and 1, inclusive.
2. The estimate of the standard error depends on the estimate of the mean (i.e., $\beta$), through $se(\beta) = \sqrt{\frac{\beta (1 - \beta)}{m + u}}$. Taking the derivative of this, we see that the maximum standard error, $\sqrt(\frac{0.25}{m + u})$, occurs at $\beta = 0.5$ and the minimum standard error, $0$, occurs at $\beta = 0, 1$.
3. We need to more than just the $\beta$-value to have a sense of how precise an estimate it is. Essentially, we need to also know the sequencing coverage of the methylation loci. Consider two CpGs, one with $m = 1, u = 3$ and the other with $m = 100$ $u = 300$. Both CpGs have $\beta = 1/4$ but the second CpG is measured with much greater precision. Assuming the binomial model, the first CpG has $se(\beta) = \sqrt{\frac{1/4 \times 3/4}{4}} = 0.22$ whereas the second CpG has $se(\beta) = \sqrt{\frac{1/4 \times 3/4}{400}} = 0.02$.

To address (2), proportion data are often transformed via a variance stabilisation transformation. The aim is to make the variance (approximately) independent of the mean. Popular variance stabilisation transformations include:

- The arcsine transformation, $\arcsin{\sqrt{\beta}}$ \citep{ANSCOMBE:1948bw}. A small value, e.g. $0.5$, is added to $m$ and $u$ to avoid $\beta = 0, 1$.
- The "averaged arcsine" transformation, $\arcsin{\sqrt{\frac{m}{m + u + 1}}} + \arcsin{\sqrt{\frac{m + 1}{m + u + 1}}}$ \citep{Freeman:1950bh}. This transformation does not have a unique inverse \citep{Nunes:2009vj}.

The use of variance stabilising transformations for proportion data seem to have fallen out of favour (e.g. [http://www.esajournals.org/doi/full/10.1890/10-0340.1](http://www.esajournals.org/doi/full/10.1890/10-0340.1)). The now favoured approach is generalised linear models (__CITE__), in particular the logistic regression model (__CITE__?).

__TODO: Cite beta-values vs. M-value paper and discuss__

@@

\subsection{Regression models}\label{sec:regression_models}

* Logistic regression
  * M-values
  * The logit transformation, $\text{logit} (\frac{\beta}{1 - \beta})$. A small value, e.g. $0.5$, is added to $m$ and $u$ to avoid $\beta = 0, 1$.
  * The probit transformation, $\Phi^{-1}(\beta)$, where $\Phi$ is the standard Normal cumulative distribution function. A small value, e.g. $0.5$, is added to $m$ and $u$ to avoid $\beta = 0, 1$.
* Beta regression

\cite{Du:2010dc} show improved inferences when using Logit transformed $\beta$-values, which they call M-values. This transformation reduces the heteroscedasticity of $\beta$-values near 0 and 1.

>>>>

@@

To begin, I consider the simple experiment of performing methylC-seq on a single sample. Once I have described the framework for a single sample, I extend it to multiple samples. The ideas here are simple, although when extended to their full generality there are several subtleties that must be kept in mind.

@@

Even with only a single sample, there are several levels of variation to consider in a bisulfite-sequencing experiment. This variation has a hierarchical structure, which I first separate into two levels, as illustrated in __CARTOON FIGURE OF METHYLC-SEQ EXPERIMENT__:

@@

$\beta$-values are the _de facto_ standard unit for reporting methylation levels due to their natural interpretation as an estimate of the average level of methylation at the locus. However, they are not necessarily the best unit for statistical inference. This is because a $\beta$-value is an estimate of a proportion and there are a well-known challenges when working with proportion data, such as:

1. The estimate of the standard error depends on the estimate of the mean (i.e., $\beta$), through $se(\beta) = \sqrt{\frac{\beta (1 - \beta)}{m + u}}$. Taking the derivative of this, we see that the maximum standard error, $\sqrt(\frac{0.25}{m + u})$, occurs at $\beta = 0.5$ and the minimum standard error, $0$, occurs at $\beta = 0, 1$.
2. We need to more than just the $\beta$-value to have a sense of how precise an estimate it is. Essentially, we need to also know the sequencing coverage of the methylation loci. Consider two CpGs, one with $m = 1, u = 3$ and the other with $m = 100$ $u = 300$. Both CpGs have $\beta = 1/4$ but the second CpG is measured with much greater precision. Assuming the binomial model, the first CpG has $se(\beta) = \sqrt{\frac{1/4 \times 3/4}{4}} = 0.22$ whereas the second CpG has $se(\beta) = \sqrt{\frac{1/4 \times 3/4}{400}} = 0.02$.
3. Proportions are bound between 0 and 1, inclusive.

To address (1), proportion data are often transformed via a variance stabilisation transformation. The aim is to make the variance independent of the mean, at least approximately. Popular variance stabilisation transformations include:

- The arcsine transformation, $\arcsin({\sqrt{\frac{m + 1}{m + u + 1}}})$ \citep{ANSCOMBE:1948bw}. A small value, in this case 1, is added to both $m$ and $u$ to avoid $\beta = 0, 1$.
- The "averaged arcsine" transformation, $\arcsin{\sqrt{\frac{m}{m + u + 1}}} + \arcsin{\sqrt{\frac{m + 1}{m + u + 1}}}$ \citep{Freeman:1950bh}. One problem with this transformation is that it does not have a unique inverse \citep{Nunes:2009vj}.

However, the general use of variance stabilising transformations for proportion data has fallen out of favour with the widespread availability of generalised linear model software, in particular for the logistic regression model \cite[e.g.,]{Warton:2011gm}.

One transformation that remains popular, at least in the analysis of DNA methylation microarray data, is the logit-transformation, also known as $\mathcal{M}$-values. An $\mathcal{M}$-value is defined as $logit_{2}(\beta) = \log_{2}(\frac{\beta}{1 - \beta} = \log_{2}\Big ( \frac{m + \alpha}{u + \alpha} \Big )$, where here $m$ and $u$ are the intensities from the methylated and unmethylated probes, respectively, and $\alpha$ is an offset to avoid a numerator or denominator that is zero. Similarly defined $\mathcal{M}$-values, also known as log-ratios, are widely and successfully used in analyses of RNA expression two-colour microarrays \cite[e.g.]{Smyth:2005ta}.

\cite{Du:2010dc} advocate for the use of $\mathcal{M}$-values in statistical analyses of methylation levels from microarray data. The main reason for this is that $\mathcal{M}$-values are approximately _homoscedastic_, i.e., their variances are approximately constant across the full range of $\mathcal{M}$-values. As already noted, the logit-transformation is not the only possible variance-stabilising transformation, but the familiarity of log-ratios to bioinformatics and genomics researchers makes it a favourable choice. Nonetheless, \cite{Du:2010dc} also advocate that the results of analyses should be reported as $\beta$-values, owing to their "more intuitive biological interpretation".

As with $\beta$-values, $\mathcal{M}$-values derived from bisulfite-sequencing data generally cannot be directly analysed due to the variable sequencing coverage across loci. However, the variable sequencing coverage can be accounted for using the afore-mentioned beta-binomial models and related regression models, which are discussed in the next section.

@@

- Don't rehash the details of Chapter WGBS analysis. Rather summarise these questions as statistical problems in the framework I outline in this chapter.
- DMC: Identify $\bar{B_{i}^{(1)}} \neq \bar{B_{i}^{(2)}}$
- DMR: Identify runs of DMCs
- DVC: Identify $\sigma(B_{i}^{(1)}) \neq \sigma(B_{i}^{(2)})$
- DVR: Identify runs of DVCs

### 2014-11-22 Rmarkdown/wgbs_analysis.Rmd
Refactor webs analysis and stat model chapters

1. Compute the M-bias from the filtered data, then call methylation events incorporating both the filters and the M-bias. This strategy requires two passes over the `SAM/BAM` file - one to compute the M-bias and one to do the methylation calling. This is the approach used by `bismark_methylation_extractor`, `comethylation` and `Bis-SNP`.
2. Call methylation events from the filtered data but retain the read-position of each methylation event. Compute the M-bias from this first file and then filter out methylation events that suffer from M-bias. This strategy requires only a single pass over the `SAM/BAM` file but requires additional information to be stored alongside the methylation calls which is then followed by a pass over the first file containing the methylation calls. This is the approach taken by `BSmooth`.

I find the first strategy conceptually simpler, and easier to program, and so use it in my methylation calling software, `comethylation`. In theory, the first pass over the `SAM/BAM` could be "skipped" by computing the M-bias simultaneously with sorting the `SAM/BAM` and marking PCR duplicates. For example, using (idealised) Unix pipes:

@@

[^cigar]: Many downstream tools, including the current version of `comethylation`, do not properly handle the information in the `CIGAR` string, particularly for soft-clipped reads. The current version of `comethylation` will skip a read (with a warning) that contains a `I` (insertion), `D` (deletion), `S` (soft-clip) or `H` (hard-clip). __FIX THIS__. This is a shortfall of the downstream tools and not of aligner-based clipping _per se_, but is nonetheless an issue in practice. __TODO: Update what changes are made to `comethylation`. The `bwa-meth` and `LAST` alignment software both perform well without pre-trimming of reads because they can soft-clip reads on the fly whereas other bisulfite-sequencing aligners, such as `Bismark`, cannot soft-clip reads and so require that reads are pre-trimmed (__CITE `bwa-meth` pre-print or paper__).__

@@

However, occasionally there are spikes in the M-bias plot, which indicate specific read-positions that we would like to exclude. __FIGURE__ shows an example of such a spike that occurred at __WHICH__ read-position, likely due to a boost in the laser for sequencing cycles $101-150$ (__CHECK: If can't find a source then rephrase__). Using `bismark_methylation_extractor` we would be forced to either retain this position or ignore read-positions $101-150$, effectively ignoring one third of the sequencing data, much of it unaffected by M-bias. My `comethylation` software is more flexible by allowing the user to exclude individual read-positions or ranges of read-positions via the `--ignore-read1-positions` and `--ignore-read2-positions` flags.

@@

[^bismark_methylation_extractor]: `Bismark` also includes a program called `bismark_methylation_extractor`, which, as the name suggests, extracts the methylation calls from the `SAM/BAM` file. So while `Bismark` performs methylation calling as part of the alignment, a secondary step using `bismark_methylation_extractor` or `comethylation` is required to make use of the methylation calls.

Both `Bismark` and `Bis-SNP` perform methylation calling of 1-tuples. In contrast, `comethylation` can perform methylation calling at m-tuples. `MethPat` (__CITE__) is also capable of performing methylation calling at m-tuples.

\subsubsection{A criticism of "identifying methylcytosines"}{sec:criticism_of_methylcytosines}

\citet{Lister:2008bh}, one of the first papers to include WGBS data, introduced an idea "to identify the presence of a methylated cytosine" [supplementary material]. This idea was also used in subsequent, and similarly high-profile, publications from the same group \citep{Lister:2009hy, Lister:2011kg}[^Lister_2013]. I believe this idea is an unrequired source of confusion, as I will explain.

@@

This is done by examining the two nucleotides upstream of the cytosine. It can be done based on the reference sequence, as is done in `Bismark` and `comethylation`, or from the reads themselves. The obvious difficulty with using the reads themselves is if the cytosine occurs at the last or second last position of the read, in which case the context may not be unambiguously determined from the the read alone. Instead, the context may be refined by initially using the reference genome context and then correcting for any sample-specific genetic variants in the two downstream bases.

@@

>>>>

1. Compute the M-bias from the filtered data, then call methylation events incorporating both the filters and the M-bias. This strategy requires two passes over the `SAM/BAM` file - one to compute the M-bias and one to do the methylation calling. This is the approach used by `bismark_methylation_extractor`, `methtuple` and `Bis-SNP`.
2. Call methylation events from the filtered data but retain the read-position of each methylation event. Compute the M-bias from this first file and then filter out methylation events that suffer from M-bias. This strategy requires only a single pass over the `SAM/BAM` file but requires additional information to be stored alongside the methylation calls which is then followed by a pass over the first file containing the methylation calls. This is the approach taken by `BSmooth`.

I find the first strategy conceptually simpler, and easier to program, and so use it in my methylation calling software, `methtuple`. In theory, the first pass over the `SAM/BAM` could be "skipped" by computing the M-bias simultaneously with sorting the `SAM/BAM` and marking PCR duplicates. For example, using (idealised) Unix pipes:

@@

[^cigar]: Many downstream tools, including the current version of `methtuple`, do not properly handle the information in the `CIGAR` string, particularly for soft-clipped reads. The current version of `methtuple` will skip a read (with a warning) that contains a `I` (insertion), `D` (deletion), `S` (soft-clip) or `H` (hard-clip). __FIX THIS__. This is a shortfall of the downstream tools and not of aligner-based clipping _per se_, but is nonetheless an issue in practice. __TODO: Update what changes are made to `methtuple`. The `bwa-meth` and `LAST` alignment software both perform well without pre-trimming of reads because they can soft-clip reads on the fly whereas other bisulfite-sequencing aligners, such as `Bismark`, cannot soft-clip reads and so require that reads are pre-trimmed (__CITE `bwa-meth` pre-print or paper__).__

@@

However, occasionally there are spikes in the M-bias plot, which indicate specific read-positions that we would like to exclude. __FIGURE__ shows an example of such a spike that occurred at __WHICH__ read-position, likely due to a boost in the laser for sequencing cycles $101-150$ (__CHECK: If can't find a source then rephrase__). Using `bismark_methylation_extractor` we would be forced to either retain this position or ignore read-positions $101-150$, effectively ignoring one third of the sequencing data, much of it unaffected by M-bias. My `methtuple` software is more flexible by allowing the user to exclude individual read-positions or ranges of read-positions via the `--ignore-read1-positions` and `--ignore-read2-positions` flags.

@@

[^bismark_methylation_extractor]: `Bismark` also includes a program called `bismark_methylation_extractor`, which, as the name suggests, extracts the methylation calls from the `SAM/BAM` file. So while `Bismark` performs methylation calling as part of the alignment, a secondary step using `bismark_methylation_extractor` or `methtuple` is required to make use of the methylation calls.

Both `Bismark` and `Bis-SNP` perform methylation calling of 1-tuples. In contrast, `methtuple` can perform methylation calling at m-tuples. `methpat` (__CITE__) is also capable of performing methylation calling at m-tuples.

@@

This is done by examining the two nucleotides upstream of the cytosine. It can be done based on the reference sequence, as is done in `Bismark` and `methtuple`, or from the reads themselves. The obvious difficulty with using the reads themselves is if the cytosine occurs at the last or second last position of the read, in which case the context may not be unambiguously determined from the the read alone. Instead, the context may be refined by initially using the reference genome context and then correcting for any sample-specific genetic variants in the two downstream bases.

@@

\subsection{A criticism of "identifying methylcytosines"}{sec:criticism_of_methylcytosines}

\citet{Lister:2008bh}, one of the first papers to include WGBS data, introduced a method "to identify the presence of a methylated cytosine" [supplementary material]. This idea was also used in subsequent, and similarly high-profile, publications from the same group \citep{Lister:2009hy, Lister:2011kg}[^Lister_2013]. I believe this idea is an unrequired source of confusion, as I will explain.
